### 2024-06-02 thesis/input_files/probabilistic.tex
June 2nd 2024

% \section{Kernel methods}
\section{Kernel quantile regression}
The idea of quantile regression has been extended to kernel methods by Takeuchi et al. \cite{takeuchi2006nonparametric}.
There, they minimise a risk functional plus regulariser defined as follows.
\begin{equation}\label{eq:kqr_min1}
    Risk[f]:=\frac{1}{m}\sum\limits_{i=1}^{m}\rho_q(y_i-f(x_i))+\frac{\tau}{2}\|w\|_\mathcal{H}^2
\end{equation}
where $f=w+b$, $w \in \mathcal{H}$ and $b \in \mathbb{R}$.
Using the link between RKHS and feature spaces, we can rewrite $f(x)=\langle w, \phi(x) \rangle_{\mathcal{H}}+b$. 
% Moreover, note that the RKHS norm is defined as follows $\|f\|_{\mathcal{H}}=\inf\{\|w\|_{\mathcal{F}}:w\in \mathcal{F},f(x)=\langle w,\varphi (x)\rangle _{\mathcal{F}},\forall x\in X\}$, where $\mathcal{F}$ is the feature space.
 Doing so we obtain a minimisation problem equivalent to minimising equation \ref{eq:kqr_min1}.
\begin{equation}\label{eq:kqr_min2}
    \begin{split}
        \min_{w,b} \quad & C \sum \limits_{i=1}^{m} \hbox{$q(y_i - \langle w, \phi(x_i) \rangle_{\mathcal{H}} - b)$} \\
        & + \hbox{$(1-q)(-y_i + \langle w, \phi(x_i) \rangle_{\mathcal{H}} + b)$} \\
        & + \frac{1}{2}\|w\|_{\mathcal{H}}^2
    \end{split}
    \end{equation}
Note the division by $\tau$ so that $C:=\frac{1}{\tau m}$.
\\
We can next rephrase the optimisation in Equation \ref{eq:kqr_min2} by introducing the slack variables $\xi_i$ and $\xi_i^*$.
\begin{equation}\label{eq:kqr_min3}
    \begin{aligned}
        \min_{w,b,\xi_i,\xi_i^*} \quad & C \sum \limits_{i=1}^{m}
        q \xi_i+ (1-q)\xi_i^*+ \frac{1}{2}\|w\|_{\mathcal{H}}^2\\
    \textrm{s.t.} \quad & y_i-\langle w, \phi(x_i) \rangle_{\mathcal{H}}-b \leq \xi_i\\
    & -y_i+\langle w, \phi(x_i) \rangle_{\mathcal{H}}+b \leq \xi_i^*\\
      &\xi_i\geq0    \\
      &\xi_i^*\geq0    \\
    \end{aligned}
    \end{equation}
In order to make it more compact, we rewrite Equation \ref{eq:kqr_min3} in matrix notation.
\begin{equation}\label{eq:kqr_min4}
    \begin{aligned}
        \min_{w,b\xi,\xi^*} \quad & C q \xi^\intercal \mathbb{1}+ C (1-q) (\xi^*)^\intercal \mathbb{1}+ \frac{1}{2}w^\intercal w\\
    \textrm{s.t.} \quad & y-\Phi^\intercal w -b \preceq \xi\\
    & -y+\Phi^\intercal w +b \preceq \xi^*\\
      &\xi\succeq0    \\
      &\xi^*\succeq0    \\
    \end{aligned}
    \end{equation}
Consider now, the Lagrangian $\mathcal{L}$ associated to Equation \ref{eq:kqr_min4}
\begin{equation}\label{eq:kqr_min5}
    \begin{split}
        L(w, b, \xi, \xi^*) = \quad & \hbox{$C q \xi^\intercal \mathbb{1} + C (1-q) (\xi^*)^\intercal \mathbb{1} + \frac{1}{2} w^\intercal w$} \\
        & \hbox{$- \lambda^\intercal (\xi - y + \Phi^\intercal w + b)$} \\
        & \hbox{$- (\lambda^*)^\intercal (\xi^* + y - \Phi^\intercal w - b)$} \\
        & \hbox{$- \nu^\intercal \xi - (\nu^*)^\intercal \xi^*$}
    \end{split}
\end{equation}
The next step is deriving its dual formulation, since it is easier and more efficient to solve. This because the dual problem has the useful property of being always convex.

@@

% In our implementatiton we take the point that minsimizes the distance between the boundary of domain of alpha, in this way we are guaranteed we do not take ith index of alpha_i for retrieving b, because it that case it would not work.
%contribution is implementation kernel quantile regression for pythos users since there exists only r version
\subsection{Weather quantiles}
In order to get acquainted with the inner workings of the presented methods, this section covers an application explaining practical details and comparing results.
\\
The dataset used is the Melbourne daily maximum temperatures \citeW{hdrcde}. It contains the daily maximum temperatures in Melbourne, Australia, from 1981-1990, excluding leap days, see Figure \ref{fig:melbourne_temperature_data}.
Due to the bimodality of the data, such dataset is commonly used to give a difficult quantile regression problem \cite{hyndman1996estimating}, thus why we chose it. The observed bimodality is that a hot day is likely to be followed by either an equally hot day or one much cooler.
\begin{figure}[!h]
    \includegraphics[width=\textwidth]{images/melbourne_temperature.png}
    \caption{Melbourne temperatures dataset}
    \label{fig:melbourne_temperature_data}
\end{figure}
Hereafter, the results of the four presented methods on the Melbourne dataset are reported, see Figure \ref{fig:melbourne_quantiles_comparison} for a visualisation. Notice, the kernel considered here is the Gaussian RBF. Hyperparameters have been tuned through cross validation, see Appendix \ref{appendix:cross_validation}.
\begin{figure}[!h]
    \begin{subfigure}[b]{0.5\linewidth}
      \centering
      \includegraphics[width=1.1\textwidth]{images/melbourne_linear_quantile_regression.png} 
      \caption{Linear quantile regression} 
      \label{fig:melbourne_linear_quantile_regression} 
      \vspace{4ex}
    \end{subfigure}%% 
    \begin{subfigure}[b]{0.5\linewidth}
      \centering
      \includegraphics[width=1.1\textwidth]{images/melbourne_gradient_boosting_quantile_regression.png} 
      \caption{Gradient boosting quantile regression} 
      \label{fig:melbourne_gradient_boosting_quantile_regression} 
      \vspace{4ex}
    \end{subfigure} 
    \begin{subfigure}[b]{0.5\linewidth}
      \centering
      \includegraphics[width=1.1\textwidth]{images/melbourne_quantile_forest.png} 
      \caption{Quantile forest} 
      \label{fig:melbourne_quantile_forest} 
    \end{subfigure}%%
    \begin{subfigure}[b]{0.5\linewidth}
      \centering
      \includegraphics[width=1.1\textwidth]{images/melbourne_gaussian_rbf_kernel_quantile_regression.png}
      \caption{Kernel quantile regression, Gaussian RBF} 
      \label{fig:melborune_kernel_quantile_regression} 
    \end{subfigure} 
    \caption{Quantile regressors comparison Melbourne weather data}
    \label{fig:melbourne_quantiles_comparison} 
  \end{figure}

% tables
\begin{table}[!h]
\caption{Pinball loss of Melbourne data}
\begin{tabular}{lllll}
    \toprule
     & Linear qr & Gbm qr & Quantile forest & KQR \\
    \midrule
    Pinball loss & 11.278895 & 10.317612 & 10.340842 & \textbf{10.031546} \\
    \bottomrule
    \end{tabular}
\end{table}

\begin{table}[!h]
    \caption{Pinball loss quantile-wise of Melbourne data}
    \begin{tabular}{lllll}
    \toprule
    Quantiles & Linear qr & Gbm qr & Quantile forest & KQR \\
    \midrule
    0.100000 & 0.710644 & 0.549232 & 0.562888 & \textbf{0.540145} \\
    0.200000 & 1.155014 & 0.938561 & 0.946712 & \textbf{0.903193} \\
    0.300000 & 1.417805 & 1.212671 & 1.222407 & \textbf{1.173627} \\
    0.400000 & 1.540108 & 1.399925 & 1.409293 & \textbf{1.368022} \\
    0.500000 & 1.574957 & 1.517281 & 1.484589 & \textbf{1.456053} \\
    0.600000 & 1.525114 & 1.498608 & 1.495474 & \textbf{1.447470} \\
    0.700000 & 1.397918 & 1.372183 & 1.362173 & \textbf{1.331375} \\
    0.800000 & 1.170140 & 1.115077 & 1.123195 & \textbf{1.096649} \\
    0.900000 & 0.787195 & \textbf{0.714075} & 0.734112 & 0.715013 \\
    \bottomrule
    \end{tabular}
\end{table}
As already pointed out, the quantile regression with $q=0.5$ corresponds to the standard regression problem, hence we can compare the proposed methods also in terms of the mean absolute error.
\begin{table}[!h]
\caption{Mean absolute error of Melbourne data}
\begin{tabular}{lllll}
    \toprule
     & Linear qr & Gbm qr & Quantile forest & KQR \\
    \midrule
    MAE & 3.253882 & 3.134805 & 3.095041 & \textbf{3.024305} \\
    \bottomrule
    \end{tabular}
\end{table}  
From these tables, we can see that kernel quantile regression outperformes the simple quantile regressor as well as the more complex models like quantile forest and gradient boosting quantile regression for the Melbourne temperatures dataset. Not only kernel quantile regression was the best in terms of total pinball loss but also the best in terms of each quantile pinball loss and mean absolute error.
Comparison has been carried out on further datasets, yielding similar conclusions as the one of above, to know more see Appendix \ref{appendix:quantile_regressor_extensive_comparison}.

@@

\begin{table}[!h]
    \caption{Kernels comparison mean absolute error Melbourne data}
    \label{tab:kernel mae comparison}
    \begin{center}
    \begin{tabular}{lll}
        \toprule
         & Kernel & MAE \\
        \midrule
        & Absolute Laplacian &  3.061553 \\
        & Matern 0.5/Laplacian &  3.061553 \\
        & Matern 1.5 &  3.046474 \\
        & Matern 2.5 &  3.034608 \\
        & Matern $\infty$/Gauss RBF &  3.024305 \\
        & Chi squared &    3.027183    \\
        & Linear &     3.142352\\
        & Periodic  & 3.118420 \\
        & Polynomial &     \textbf{3.022743} \\
        & Sigmoid &      3.108107       \\
        & Cosine &    4.849589 \\
        \bottomrule
        \end{tabular}
    \end{center}
    \end{table}

>>>>

% \section{Kernel methods}
\section{Kernel quantile regression}
The idea of quantile regression has been extended to kernel methods by Takeuchi et al.\ \cite{takeuchi2006nonparametric}.
There, they minimise a risk functional plus regulariser defined as follows.
\begin{equation}\label{eq:kqr_min1}
    Risk[f]:=\frac{1}{m}\sum\limits_{i=1}^{m}\rho_q(y_i-f(x_i))+\frac{\tau}{2}\|w\|_\mathcal{H}^2
\end{equation}
where $f=w+b$, $w \in \mathcal{H}$ and $b \in \mathbb{R}$.
Using the link between RKHS and feature spaces, we can rewrite $f(x)=\langle w, \phi(x) \rangle_{\mathcal{H}}+b$. 
% Moreover, note that the RKHS norm is defined as follows $\|f\|_{\mathcal{H}}=\inf\{\|w\|_{\mathcal{F}}:w\in \mathcal{F},f(x)=\langle w,\varphi (x)\rangle _{\mathcal{F}},\forall x\in X\}$, where $\mathcal{F}$ is the feature space.
 Doing so we obtain a minimisation problem equivalent to minimising equation \ref{eq:kqr_min1}.
\begin{equation}\label{eq:kqr_min2}
    \begin{split}
        \min_{w,b} \quad & C \sum \limits_{i=1}^{m} \hbox{$q(y_i - \langle w, \phi(x_i) \rangle_{\mathcal{H}} - b)$} \\
        & + \hbox{$(1-q)(-y_i + \langle w, \phi(x_i) \rangle_{\mathcal{H}} + b)$} \\
        & + \frac{1}{2}\|w\|_{\mathcal{H}}^2
    \end{split}
    \end{equation}
Note the division by $\tau$ so that $C:=\frac{1}{\tau m}$.
\\
We can next rephrase the optimisation in Equation \ref{eq:kqr_min2} by introducing the slack variables $\xi_i$ and $\xi_i^*$.
\begin{equation}\label{eq:kqr_min3}
    \begin{aligned}
        \min_{w,b,\xi_i,\xi_i^*} \quad & C \sum \limits_{i=1}^{m}
        q \xi_i+ (1-q)\xi_i^*+ \frac{1}{2}\|w\|_{\mathcal{H}}^2\\
    \textrm{s.t.} \quad & y_i-\langle w, \phi(x_i) \rangle_{\mathcal{H}}-b \leq \xi_i\\
    & -y_i+\langle w, \phi(x_i) \rangle_{\mathcal{H}}+b \leq \xi_i^*\\
      &\xi_i\geq0    \\
      &\xi_i^*\geq0    \\
    \end{aligned}
    \end{equation}
In order to make it more compact, we rewrite Equation \ref{eq:kqr_min3} in matrix notation.
\begin{equation}\label{eq:kqr_min4}
    \begin{aligned}
        \min_{w,b\xi,\xi^*} \quad & C q \xi^\intercal \mathbb{1}+ C (1-q) (\xi^*)^\intercal \mathbb{1}+ \frac{1}{2}w^\intercal w\\
    \textrm{s.t.} \quad & y-\Phi^\intercal w -b \preceq \xi\\
    & -y+\Phi^\intercal w +b \preceq \xi^*\\
      &\xi\succeq0    \\
      &\xi^*\succeq0    \\
    \end{aligned}
    \end{equation}
Consider now, the Lagrangian $\mathcal{L}$ associated to Equation \ref{eq:kqr_min4}
\begin{equation}\label{eq:kqr_min5}
    \begin{split}
        L(w, b, \xi, \xi^*) = \quad & \hbox{$C q \xi^\intercal \mathbb{1} + C (1-q) (\xi^*)^\intercal \mathbb{1} + \frac{1}{2} w^\intercal w$} \\
        & \hbox{$- \lambda^\intercal (\xi - y + \Phi^\intercal w + b)$} \\
        & \hbox{$- (\lambda^*)^\intercal (\xi^* + y - \Phi^\intercal w - b)$} \\
        & \hbox{$- \nu^\intercal \xi - (\nu^*)^\intercal \xi^*$}
    \end{split}
\end{equation}
The next step is deriving its dual formulation, since it is easier and more efficient to solve. This because the dual problem has the useful property of being always convex.

@@

% In our implementatiton we take the point that minsimizes the distance between the boundary of domain of alpha, in this way we are guaranteed we do not take ith index of alpha_i for retrieving b, because it that case it would not work.
%contribution is implementation kernel quantile regression for pythos users since there exists only r version
\subsection{Weather quantiles}
In order to get acquainted with the inner workings of the presented methods, this section covers an application explaining practical details and comparing results.
\\
The dataset used is the Melbourne daily maximum temperatures \citeW{hdrcde}. It contains the daily maximum temperatures in Melbourne, Australia, from 1981-1990, excluding leap days, see Figure \ref{fig:melbourne_temperature_data}.
Due to the bimodality of the data, such dataset is commonly used to give a difficult quantile regression problem \cite{hyndman1996estimating}, thus why we chose it. The observed bimodality is that a hot day is likely to be followed by either an equally hot day or one much cooler.
\begin{figure}[!h]
    \includegraphics[width=\textwidth]{images/melbourne_temperature.png}
    \caption{Melbourne temperatures dataset}
    \label{fig:melbourne_temperature_data}
\end{figure}
Hereafter, the results of the four presented methods on the Melbourne dataset are reported, see Figure \ref{fig:melbourne_quantiles_comparison} for a visualisation. Notice, the kernel considered here is the Gaussian RBF. Hyperparameters have been tuned through cross validation, see Appendix \ref{appendix:cross_validation}.
\begin{figure}[!h]
    \begin{subfigure}[b]{0.5\linewidth}
      \centering
      \includegraphics[width=1.1\textwidth]{images/melbourne_linear_quantile_regression.png} 
      \caption{Linear quantile regression} 
      \label{fig:melbourne_linear_quantile_regression} 
      \vspace{4ex}
    \end{subfigure}%% 
    \begin{subfigure}[b]{0.5\linewidth}
      \centering
      \includegraphics[width=1.1\textwidth]{images/melbourne_gradient_boosting_quantile_regression.png} 
      \caption{Gradient boosting quantile regression} 
      \label{fig:melbourne_gradient_boosting_quantile_regression} 
      \vspace{4ex}
    \end{subfigure} 
    \begin{subfigure}[b]{0.5\linewidth}
      \centering
      \includegraphics[width=1.1\textwidth]{images/melbourne_quantile_forest.png} 
      \caption{Quantile forest} 
      \label{fig:melbourne_quantile_forest} 
    \end{subfigure}%%
    \begin{subfigure}[b]{0.5\linewidth}
      \centering
      \includegraphics[width=1.1\textwidth]{images/melbourne_gaussian_rbf_kernel_quantile_regression.png}
      \caption{Kernel quantile regression, Gaussian RBF} 
      \label{fig:melborune_kernel_quantile_regression} 
    \end{subfigure} 
    \caption{Quantile regressors comparison for Melbourne weather data}
    \label{fig:melbourne_quantiles_comparison} 
  \end{figure}

% tables
\begin{table}[!h]
\caption{Pinball loss for Melbourne data}
\begin{tabular}{lllll}
    \toprule
     & Linear qr & Gbm qr & Quantile forest & KQR \\
    \midrule
    Pinball loss & 11.278895 & 10.317612 & 10.340842 & \textbf{10.031546} \\
    \bottomrule
    \end{tabular}
\end{table}

\begin{table}[!h]
    \caption{Pinball loss quantile-wise for Melbourne data}
    \begin{tabular}{lllll}
    \toprule
    Quantiles & Linear qr & Gbm qr & Quantile forest & KQR \\
    \midrule
    0.100000 & 0.710644 & 0.549232 & 0.562888 & \textbf{0.540145} \\
    0.200000 & 1.155014 & 0.938561 & 0.946712 & \textbf{0.903193} \\
    0.300000 & 1.417805 & 1.212671 & 1.222407 & \textbf{1.173627} \\
    0.400000 & 1.540108 & 1.399925 & 1.409293 & \textbf{1.368022} \\
    0.500000 & 1.574957 & 1.517281 & 1.484589 & \textbf{1.456053} \\
    0.600000 & 1.525114 & 1.498608 & 1.495474 & \textbf{1.447470} \\
    0.700000 & 1.397918 & 1.372183 & 1.362173 & \textbf{1.331375} \\
    0.800000 & 1.170140 & 1.115077 & 1.123195 & \textbf{1.096649} \\
    0.900000 & 0.787195 & \textbf{0.714075} & 0.734112 & 0.715013 \\
    \bottomrule
    \end{tabular}
\end{table}
As already pointed out, the quantile regression with $q=0.5$ corresponds to the standard regression problem, hence we can compare the proposed methods also in terms of the mean absolute error.
\begin{table}[!h]
\caption{Mean absolute error for Melbourne data}
\begin{tabular}{lllll}
    \toprule
     & Linear qr & Gbm qr & Quantile forest & KQR \\
    \midrule
    MAE & 3.253882 & 3.134805 & 3.095041 & \textbf{3.024305} \\
    \bottomrule
    \end{tabular}
\end{table}  
From these tables, we can see that kernel quantile regression outperformes the simple quantile regressor as well as the more complex models like quantile forest and gradient boosting quantile regression for the Melbourne temperatures dataset. Not only kernel quantile regression was the best in terms of total pinball loss but also the best in terms of each quantile pinball loss and mean absolute error.
Comparison has been carried out on further datasets, yielding similar conclusions as the one of above, to know more see Appendix \ref{appendix:quantile_regressor_extensive_comparison}.

@@

\begin{table}[!h]
    \caption{Kernels comparison mean absolute error for Melbourne data}
    \label{tab:kernel mae comparison}
    \begin{center}
    \begin{tabular}{lll}
        \toprule
         & Kernel & MAE \\
        \midrule
        & Absolute Laplacian &  3.061553 \\
        & Matern 0.5/Laplacian &  3.061553 \\
        & Matern 1.5 &  3.046474 \\
        & Matern 2.5 &  3.034608 \\
        & Matern $\infty$/Gauss RBF &  3.024305 \\
        & Chi squared &    3.027183    \\
        & Linear &     3.142352\\
        & Periodic  & 3.118420 \\
        & Polynomial &     \textbf{3.022743} \\
        & Sigmoid &      3.108107       \\
        & Cosine &    4.849589 \\
        \bottomrule
        \end{tabular}
    \end{center}
    \end{table}

### 2024-06-02 thesis/input_files/problemdescription.tex
June 2nd 2024

Individuals and organizations constantly face situations of uncertainty, thus the need for robust forecasting methods. Such methods are crucial in the process of taking informed decisions and for strategic planning.
\\
The basic idea of forecasting is that we can extract knowledge from the past in order to make educated guesses about the future. Consequently, the range of fields where forecasting can be applied is very wide.
In this thesis, our focus lies on applying forecasting techniques to the energy sector. 
\\
Our decision to focus on the energy market is mainly motivated by the rapid changes it has undergone. %/experienced 
Over the last decades, electricity markets have gone through an unprecedented transformation. This shift was driven by the liberalization of such markets, the development and integration of renewable energy sources, the  increase of low carbon technologies and the adoption of smart meters. Events like the California electricity crisis are further motivating the choice of the electricity sector as subject of our studies, see \cite{california}.
Moreover, the process of deregulation lead to an increasing interest in the field of electricity forecasting (EF) within the academic community, see Figure \ref{fig:epf_evolution}.
In addtion, the United Nations have identified the right to access affordable, reliable, sustainable and modern energy as one of their 17 sustainable development goals (SDGs) \citeW{un_sdgs}.
Finally, the electricity market has a set of features that make it unique: electricity cannot be stored in an efficient way and supply and demand have to be matched instantly.
\\
\section{Motivation}
There are mutliple reasons why the energy sector needs robust forecasting techniques.
For power market companies, being able to predict prices with a low mean absolute percentage error (MAPE) \ref{mape} results in increased savings \cite{savings}. Furthermore, the adoption of smart meters provides power market companies with a huge amount of consumer data. This enables them to better model consumer preferences.
\\
Transmission system operators' (TSO) main goal is to match supply and demand, generally TSO do so by increasing or decreasing the generation. Thus, from their point of view forecasting is critical for balancing the electricity network.
Probabilistic forecasting may be useful to power producers, traders and consumers in order to improve their decision making process and managing risk. This holds in particular for traders, because probabilistic forecasts enable them to simulate scenarios and carry out stress tests.
Other possible applications include: control of storage, demand side response, anomaly detection, network design and planning, simulating inputs and handling missing data.
\\
\section{Point versus probabilistic forecast}
A distinction has to be made between two types of forecasting approaches: point forecasts and probabilistic forecasts.
Point prediction, also called deterministic forecasting in the literature \cite{EPF_review}, is all about predicting a particular value in time.
On the other hand, with probabilistic forecasting we aim at predicting either an interval, quantiles or a probability distribution for each point in time \cite{nowotarski}. For this reason, probabilistic forecasts are more informative than point forecasts. This is why the interest of the research community is shifting towards them.
A probabilistic forecast can be turned into a point forecast by simply taking its expectation.
Alternatively, a probabilistic forecast can be derived from a point one by modeling the residuals of the point prediction.

>>>>

Individuals and organisations constantly face situations of uncertainty, thus the need for robust forecasting methods. Such methods are crucial in the process of taking informed decisions and for strategic planning.
\\
The basic idea of forecasting is that we can extract knowledge from the past in order to make educated guesses about the future. Consequently, the range of fields where forecasting can be applied is very wide.
In this thesis, our focus lies on applying forecasting techniques to the energy sector. 
\\
Our decision to focus on the energy market is mainly motivated by the rapid changes it has undergone. %/experienced 
Over the last decades, electricity markets have gone through an unprecedented transformation. This shift was driven by the liberalisation of such markets, the development and integration of renewable energy sources, the  increase of low carbon technologies and the adoption of smart meters. Events like the California electricity crisis are further motivating the choice of the electricity sector as subject of our studies, see \cite{california}.
Moreover, the process of deregulation lead to an increasing interest in the field of electricity forecasting (EF) within the academic community, see Figure \ref{fig:epf_evolution}.
In addtion, the United Nations have identified the right to access affordable, reliable, sustainable and modern energy as one of their 17 sustainable development goals (SDGs) \citeW{un_sdgs}.
Finally, the electricity market has a set of features that make it unique: electricity cannot be stored in an efficient way and supply and demand have to be matched instantly.
\\
\section{Motivation}
There are mutliple reasons why the energy sector needs robust forecasting techniques.
For power market companies, being able to predict prices with a low mean absolute percentage error (MAPE) \ref{mape} results in increased savings \cite{savings}. Furthermore, the adoption of smart meters provides power market companies with a huge amount of consumer data. This enables them to better model consumer preferences.
\\
Transmission system operators' (TSO) main goal is to match supply and demand, generally TSO do so by increasing or decreasing the generation. Thus, from their point of view forecasting is critical for balancing the electricity network.
Probabilistic forecasting may be useful to power producers, traders and consumers in order to improve their decision making process and managing risk. This holds in particular for traders, because probabilistic forecasts enable them to simulate scenarios and carry out stress tests.
Other possible applications include: control of storage, demand side response, anomaly detection, network design and planning, simulating inputs and handling missing data.
\\
\section{Point versus probabilistic forecast}
A distinction has to be made between two types of forecasting approaches: point forecasts and probabilistic forecasts.
Point prediction, also called deterministic forecasting in the literature \cite{EPF_review}, is all about predicting a particular value in time.
On the other hand, with probabilistic forecasting we aim at predicting either an interval, quantiles or a probability distribution for each point in time \cite{nowotarski}. For this reason, probabilistic forecasts are more informative than point forecasts. This is why the interest of the research community is shifting towards them.
A probabilistic forecast can be turned into a point forecast by simply taking its expectation.
Alternatively, a probabilistic forecast can be derived from a point one by modeling the residuals of the point prediction.

### 2024-06-02 thesis/input_files/qr.tex
June 2nd 2024

Quantile regression can be interpreted as an extension of standard regression. In this setting, one slices the dependent variables into quantiles and then fit a regression for each quantile. With standard regression, we build a model for the conditional mean, conversely, with quantile regression we model the conditional quantile function for any desired quantile. 
Therefore, with quantile regression we are able to study the impact of covariates on quantiles directly.
\begin{definition}
    For any real valued random variable Y, we define its associated quantile function.
    \begin{equation}
        Q_q=\inf\{y:F(y)\geq q\}
    \end{equation}
\end{definition}
Alternatively, in order  to ease the posing of the quantile regression problem, we can formulate quantiles as the solutions to the following optimisation problems.
\\
For any $0<q<1$ consider the pinball loss function from section \ref{pinball}, $\rho_q(u)=u(q-\mathbb{I}_{\{u<0\}})$. 
Such loss is minimized by the quantile $Q_q$.
 we can estimate quantiles by minimizing the expectation of $\rho_q(y-g(x,\beta))$ with respect to the parameter $\beta$.
\\
Note, that in the special case $q=\frac{1}{2}$,  quantile regression corresponds exactly to standard regression with an absolute value loss function.
\\
It follows that the conditional linear quantile function $\hat{Q}_q=x_i\beta(q)$, can be estimated by solving
\begin{equation}\label{eq:linear_quantile_regression_minimisation_problem}
    \hat{\beta}(q)=\argmin{\beta}\sum\limits_{i=1}^n \rho_q   (y_i-x_i \beta)
\end{equation}
\\
Notice that, this cost function is not differentiable, therefore there is no analytical solution to the quantile regression problem. Nevertheless, we can easily solve it by employing linear programming and convex optimisation \cite{boyd2004convex}.
\\
Furthermore, we can extend this framework to non linear quantile regression by choosing a non linear model in place of $x\beta$ in the above Equation \ref{eq:linear_quantile_regression_minimisation_problem}.
% \\
% In order to illustrate the approach, let us consider applying quantile regression to the maximum daily temperature dataset in Melbourne \cite{hyndman1996estimating}; models considered will be linear, gradient boosting machine, quantile forest\cite{meinshausen2006quantile} and kernel quantile regression.
% - se si puo fare plot con stessi dati ma differenti f nel quantile regression

>>>>

Quantile regression can be interpreted as an extension of standard regression. In this setting, one slices the dependent variables into quantiles and then fits a regression for each quantile. With standard regression, we build a model for the conditional mean, conversely, with quantile regression we model the conditional quantile function for any desired quantile. 
Therefore, with quantile regression we are able to study the impact of covariates on quantiles directly.
\begin{definition}
    For any real valued random variable Y, we define its associated quantile function.
    \begin{equation}
        Q_q=\inf\{y:F(y)\geq q\}
    \end{equation}
\end{definition}
Alternatively, in order  to ease the posing of the quantile regression problem, we can formulate quantiles as the solutions to the following optimisation problems.
\\
For any $0<q<1$ consider the pinball loss function from Section \ref{pinball}, $\rho_q(u)=u(q-\mathbb{I}_{\{u<0\}})$. 
Such loss is minimized by the quantile $Q_q$.
 we can estimate quantiles by minimizing the expectation of $\rho_q(y-g(x,\beta))$ with respect to the parameter $\beta$.
\\
Note, that in the special case $q=\frac{1}{2}$,  quantile regression corresponds exactly to standard regression with an absolute value loss function.
\\
It follows that the conditional linear quantile function $\hat{Q}_q=x_i\beta(q)$, can be estimated by solving
\begin{equation}\label{eq:linear_quantile_regression_minimisation_problem}
    \hat{\beta}(q)=\argmin{\beta}\sum\limits_{i=1}^n \rho_q   (y_i-x_i \beta)
\end{equation}
\\
Notice that, this cost function is not differentiable, therefore there is no analytical solution to the quantile regression problem. Nevertheless, we can easily solve it by employing linear programming and convex optimisation \cite{boyd2004convex}.
\\
Furthermore, we can extend this framework to non linear quantile regression by choosing a non linear model in place of $x\beta$ in the above Equation \ref{eq:linear_quantile_regression_minimisation_problem}.
% \\
% In order to illustrate the approach, let us consider applying quantile regression to the maximum daily temperature dataset in Melbourne \cite{hyndman1996estimating}; models considered will be linear, gradient boosting machine, quantile forest\cite{meinshausen2006quantile} and kernel quantile regression.
% - se si puo fare plot con stessi dati ma differenti f nel quantile regression
